A Mechanised Library for Generic (Co)Programming

### 2024-03-18 paper/paper.tex
minor

\documentclass[anonymous, a4paper, UKenglish, cleveref, autoref, thm-restate]{lipics-v2021}
\usepackage[utf8]{inputenc} % Required for inputting international characters
\usepackage[T1]{fontenc} % Output font encoding for international characters
% \usepackage[backend=bibtex,style=alphabetic,natbib=true]{biblatex}
% \usepackage[backend=bibtex,style=alphabetic,natbib=true]{biblatex}
% \addbibresource{references.bib}
\usepackage{amsmath, amssymb, amsfonts, amsthm, stmaryrd}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{graphicx}
\usepackage{enumitem}
\usepackage{mathpartir}
\usepackage{bussproofs}
\usepackage{xparse}
\usepackage[usenames, dvipsnames]{xcolor}
\usepackage{lipsum}
\usepackage{xargs}
\usepackage{hyperref}
\usepackage{tikz-cd}
\usepackage{todonotes}
\usepackage{url}
\usepackage{xspace}
\usepackage{rotating}
\usepackage{quiver}
\usepackage{minted}
\usepackage{newunicodechar}
\usepackage{microtype}
% \addbibresource{references.bib}
\usepackage{array}   % for \newcolumntype macro
\newcolumntype{C}{>{$}c<{$}} % math-mode version of "l" column type
\newcolumntype{L}{>{$}l<{$}} % math-mode version of "l" column type

@@

\section{Introduction}
\label{sec:intro}
Recursive definitions cannot be proven well-defined automatically due to the
halting problem. Modern proof assistants like Coq or Agda provide a sound, but
incomplete algorithm which syntactically checks for termination or
productivity.  For recursion, this is done by automatically inferring
which argument in the recursive call gets smaller with respects to the original
input argument.  For productivity, the algorithm checks that the corecursive
call appears directly under a constructor to make sure that this function
always produces at least one element after each recursive step.
This implies that some functions, though well-defined, cannot be accepted by
the proof assistant. Examples of this include common sorting algorithms, such
as Quicksort and Mergesort.
%One example is Quicksort, written in OCaml
%below:
\begin{minted}{OCaml}
let rec qsort xs = match divide xs with | None -> []
 | Some (pivot, (smaller, larger)) -> qsort smaller @ (pivot::qsort larger)
\end{minted}
While \ocaml{qsort} is a well-defined mathematical function it cannot be
accepted by a proof assistant.  The reason is that the \ocaml{divide}
function \emph{destructuring} the input ``dives'' deeper in the input returning
two sublists with the head as a pivot.

@@

This work presents the first formalisation in Coq of \emph{hylomorphisms}, as
well as their algebraic laws that: (1) is \textbf{fully axiom-free}; and (2)
allows the extraction of naturally-looking, clean OCaml code. While programmers
still need to reason about the termination of their programs, our mechanisation
will allow the use of the algebraic laws of hylomorphisms for program
reasoning, as well as extracting to clean, naturally-looking code. As an
example our mechanisation of \ocaml{qsort} as a hylomorphism will produce
the following code:
\begin{coqcode}
let rec qsort = function | [] -> []
  | h :: t -> let (l, r) = partition (fun x0 -> (<=) x0 h) t in
              let cx = fun p -> match p with | Lbranch -> l | Rbranch -> r in 
              app (qsort (cx Lbranch)) (h :: (qsort (cx Rbranch)))
\end{coqcode}
The contributions we make in this paper are as follows:
\begin{itemize}
  \item We provide a framework for \emph{generic programming} with recursion
    schemes in Coq, with a proof of their algebraic laws and 
    program equivalences, that allows clean code extraction.
  \item We formalise divide and conquer, dynamic programming, and
    mutual recursion algorithms.
  \item We verified the short-cut fusion optimisation in Coq and
    extract optimised code to OCaml.
\end{itemize}

@@

\dcas{Marco, shouldn't the below be "given an F-algebra and a recursive
F-coalgebra"? We contradict ourselves below where we say that the hylomorphism
equation does not necessarily have unique solutions, or solutions at all. As
far as I understand, these are also called hylomorphisms. E.g. one can define a
general fixpoint combinator as a hylomorphism, which does not have a unique
solution}
Thus, in general given an $F$-algebra and $F$-coalgebra, the hylomorphism is the
unique solution to the equation
\begin{equation}
  \label{eq:hylo}
  f = a \circ F\; f \circ c
\end{equation}
However this solution does not necessarily exist for an arbitrary pair of
algebras and coalgebras. In fact, this definition cannot be accepted by Coq,
since there is no structurally smaller argument to do recursion on and there is
no corecursive definition that is guaranteed to be syntactically guarded -- as
required by Coq -- due to the use of an arbitrary algebra $a$. Furthermore,
solutions to the hylomorphism equation are \emph{not} guaranteed to be unique,
or to even exist. To address these issues, we mechanise \emph{recursive
hylomorphisms} which are guaranteed to have a unique solution to the
hylomorphism equation. These are hylomoprhisms where the coalgebra is
\emph{recursive}, i.e. coalgebras that terminate on all inputs. The following
predicate captures that a coalgebra terminates on an input: 
\begin{coqcode}
Inductive RecF (h : Coalg F A) : A -> Prop :=
| RecF_fold x : (forall e, RecF h (cont (h x) e)) -> RecF h x.
\end{coqcode}
A recursive coalgebra of type \coq{RCoalg F A} is a coalgebra 
\coq{c : Coalg F A} such that \coq{forall x, RecF c x}.
Recursive hylomorphisms are implemented in Coq recursively on the structure of
the proof for the type (\coq{RecF}) as follows:
\begin{coqcode}
Definition hylo_def (a : Alg F B) (c : Coalg F A) 
  : forall (x : A), RecF c x -> B := fix f x H
  := match c x as h0 return (forall e : Pos (shape h0), RecF c (cont h0 e)) -> B
     with | MkApp s_x c_x => fun H => a (MkApp s_x (fun e => f (c_x e) (H e)))
     end (RecF_inv H).
\end{coqcode}
We use \coq{RecF_inv} to obtain the structurally smaller proof to use in
the recursive calls. As we did with catamorphisms and anamorphisms, we
prove that \coq{hylo_def} is respectful, and use this proof to build the
corresponding higher-order proper morphism:
\begin{coqcode}
  hylo : forall `{F : Cont Sh P} `{setoid A} `{setoid B}, 
    Alg F B ~> RCoalg F A ~> A ~> B
\end{coqcode}
Finally, we show that recursive hylomorphisms are the unique solution to the
hylomorphism equation.
\begin{coqcode}
Lemma hylo_univ (g : Alg F B) (h : RCoalg F A) (f : A ~> B)
    : f =e hylo g h <-> f =e g \o fmap f \o h.
\end{coqcode}
Hylomorphisms fusion falls out from the universal property:
\begin{coqcode}
Lemma hylo_fusion_l (h1 : RCoalg F A) (g1 : Alg F B) (g2 : Alg F C)
  (f2 : B ~> C) (E2 : f2 \o g1 =e g2 \o fmap f2) : f2 \o hylo g1 h1 =e hylo g2 h1.
\end{coqcode}
% Qed.
%   rewrite hylo_univ, fmap_comp, compA, <- E2,
%           <- compA, <- compA, (compA g1), <- hylo_unr.  reflexivity.

@@

% Similarly to the rest of the definitions, we prove that recursive anamorphisms
% are respectful, and use this fact to build them as higher order proper
% morphisms, and prove their universal law, i.e.\ that recursive anamorphisms are
% the unique solution to the equation
% \coq{f =e l_in \o fmap f \o h},
% where \coq{h} is a recursive coalgebra.
\section{Extraction}
We now proceed to calculate recursive programs from simple, non-recursive specifications.
The steps are the following:
\begin{enumerate}
  \item Write a specification as a morphism \coq{f : A ~> B}.
  \item Use Coq's generalised rewriting and the laws of morphisms to 
    rewrite it to some \coq{g : A ~> B} such that \coq{f =e g}.
  \item Project the function \coq{h : A -> B} from \coq{g}
    (i.e. \coq{h = app g}).
  \item Extract \coq{h} to OCaml.
\end{enumerate}
Note that steps 2 and 3 are \emph{proofs} that \coq{f} and \coq{h} are
\emph{extensionally equal}. Then, Coq's extraction mechanism will erase any
proofs from the code, and inline our definitions. To ease the application of
these steps, we provide the following helper type:
\begin{coqcode}
  Record Ext (f : A ~> B) := 
    MkExt { target :> A -> B; tgt_eq : app f =e target; }.
\end{coqcode}
Note that \coq{tgt_eq} will be erased when extracting \coq{Ext f} to OCaml,
since it is in \coq{Prop}. This implies that the code extracted for \coq{Ext f}
will be the function \coq{target} that is extensionally equal to \coq{f}.  We
use a shortcut tactic, \coq{calculate}, that is a synonym to 
\coq{eapply MkExt}. As a result of this tactic, we will be given the goal 
\coq{f =e ?Goal0}. We can rewrite \coq{f}, unfold definitions, etc, and
complete the definition with \coq{reflexivity}, and this will instantiate the
metavariable \coq{?Goal0} to our desired function. We will use this extensively
in our examples.

\subsection{Sorting Algorithms}

So far, our examples could be modelled exclusively using catamorphisms. Here,
we show examples that require recursive hylomorphisms and termination proofs.
We complete the sorting algorithm examples by applying fusion optimisation.

\subsubsection{Binary Trees}

Both Mergesort and Quicksort are hylomorphisms whose recursive structure is
described by a binary tree. In particular, the container used in their
definition is isomorphic to \coq{TreeF (list nat) unit} and 
\coq{TreeF unit nat} respectively, where \coq{TreeF} is defined as follows:
\begin{coqcode}
Inductive ITreeF L N X := i_leaf (l : L) | i_node (n : N) (l r : X)
\end{coqcode}
The definition as a container uses the following shapes and positions:
\begin{coqcode}
Inductive Tshape L A := | Leaf (ELEM : L) | Node (ELEM : A).
Inductive Tpos := | Lbranch | Rbranch. 
\end{coqcode}
We define the tree container, \coq{TreeF}, in a straightforward way, by making
the positions of type \coq{Tpos} only valid in \coq{Node}.
We define a series of definitions for tree container constructors and 
destructors:
\begin{coqcode}
Definition a_out {L A X : Type} : App (TreeF L A) X ~> ITreeF L A X.
(* Converting TreeF to ITreeF *)
Notation a_leaf x := (MkCont (Leaf _ x) (@dom_leaf _ _ _ x)).
Notation a_node x l r := (MkCont (Node _ x)
  (fun p => match val p with | Lbranch => l | Rbranch => r end)).
\end{coqcode}

\subsubsection{Mergesort}

Mergesort is a hylomorphism with the recursive structure defined by the
container \coq{(TreeF (list int) unit)}. 
The computationally expensive part of the Mergesort implementation is the
algebra, since this is where most of the comparisons happen. The recursive
coalgebra for Mergesort is defined as the following Coq functions:
\begin{coqcode}
Fixpoint splitL (x : list int) (accL accR : list int) :=
  match x with
  | nil => (accL, accR)
  | cons x xs => splitL xs accR (cons x accL)
  end.

Definition c_split : Coalg (TreeF (list int) unit) (list int) :=
|{ x ~> match x with
        | nil | cons _ nil => a_leaf x
        | _ => let (l, r) := splitL x nil nil in
               a_node tt l r
        end
 }|.
Defined.
\end{coqcode}
Here, the notation \coq{|{ x ~> E }|} defines the morphism \coq{fun x => E},
and uses a custom tactic to prove that it is respectful. The recursive
coalgebra \coq{c_split} takes a list, and it returns a leaf with the input
list, if its size is less or equal than 1. Otherwise, it applies the function
\coq{splitL}, and divides the list in two halves.

It is easy to prove that this coalgebra is recursive, by showing that
it respects the relation \coq{<} on the length of the lists:
\begin{coqcode}
Lemma split_fin : respects_relation c_split (@length int) lt.
  Proof. (* *) Qed.

Definition tsplit : RCoalg (TreeF (list int) unit) (list int)
  := mk_wf_coalg wf_lt split_fin.
\end{coqcode}

The algebra simply applies the \coq{mergeL} function in order to the sublists
in every node:
\begin{coqcode}
Fixpoint mergeL l1 l2 {struct l1}  : list int := (* *)

Definition merge : App (TreeF (list int) unit) (list int) ~> list int.
|{ x ~> match a_out x with | node _ l r => mergeL l r | leaf e => e end }|.
Defined.
\end{coqcode}

Putting everything together, Mergesort is defined as the following
hylomorphism:
\begin{coqcode}
Definition msort := hylo merge tsplit.
\end{coqcode}
Note, however, that we could have composed the catamorphism and anamorphism,
and used the deforestation result to optimise the intermediate binary tree
away:
\begin{coqcode}
Definition msort : Ext (cata merge \o rana tsplit).
calculate. rewrite (deforest l_in_out). simpl. reflexivity. Defined.
\end{coqcode}
In both cases, the result is the same, and it extracts to the usual divide and
conquer definition of Mergesort in OCaml:
\begin{ocamlcode}
let rec msort x = match x with
| [] -> x | _ :: l -> (match l with | [] -> x | _ :: _ ->
     let (l0, r) = splitL x [] [] in
     let cx = fun p -> match p with | Lbranch -> l0 | Rbranch -> r in
     mergeL (msort (cx Lbranch)) (msort (cx Rbranch)))
\end{ocamlcode}

\subsubsection{Quicksort}

@@

Definition c_split : Coalg (TreeF unit int) (list int).
|{ x ~> match x with
        | nil => a_leaf tt
        | cons h t => let (l, r) := List.partition (fun x => x <=? h) t in
                      a_node h l r
        end}|.
Defined.
\end{coqcode}
As with the Mergesort case, we prove that the coalgebra is recursive by showing
that it respects the ``less-than'' relation on the length of the lists.
We then define, and extract, the hylomorphism \coq{hylo merge r_split}, which
extracts to the following OCaml code:
\begin{ocamlcode}
let rec qsort = function
| [] -> [] | h :: t ->
  let (l, r) = partition (fun x0 -> leb x0 h) t in
  let x0 = fun e -> qsort (match e with | Lbranch -> l | Rbranch -> r) in 
  app (x0 Lbranch) (h :: (x0 Rbranch))
\end{ocamlcode}
Note that Coq's code extraction is unable to inline \coq{x0}, but the resulting
code is similar to a hand-written \coq{qsort}.

\subsubsection{Fusing a divide-and-conquer computation}

>>>>

\documentclass[ a4paper, UKenglish, cleveref, autoref, thm-restate]{lipics-v2021}
\usepackage[utf8]{inputenc} % Required for inputting international characters
\usepackage[T1]{fontenc} % Output font encoding for international characters
% \usepackage[backend=bibtex,style=alphabetic,natbib=true]{biblatex}
% \usepackage[backend=bibtex,style=alphabetic,natbib=true]{biblatex}
% \addbibresource{references.bib}
\usepackage{amsmath, amssymb, amsfonts, amsthm, stmaryrd}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{graphicx}
\usepackage{enumitem}
\usepackage{mathpartir}
\usepackage{bussproofs}
\usepackage{xparse}
\usepackage[usenames, dvipsnames]{xcolor}
\usepackage{lipsum}
\usepackage{xargs}
\usepackage{hyperref}
\usepackage{tikz-cd}
\usepackage{todonotes}
\usepackage{url}
\usepackage{xspace}
\usepackage{rotating}
\usepackage{quiver}
\usepackage{minted}
\usepackage{newunicodechar}
\usepackage{microtype}
% \addbibresource{references.bib}
\usepackage{array}   % for \newcolumntype macro
\newcolumntype{C}{>{$}c<{$}} % math-mode version of "l" column type
\newcolumntype{L}{>{$}l<{$}} % math-mode version of "l" column type

@@

\section{Introduction}
\label{sec:intro}
Recursive definitions cannot be proven well-defined automatically due to the
halting problem. Modern proof assistants like Coq or Agda provide a sound, but
incomplete algorithm which syntactically checks for termination or
productivity.  For recursion, this is done by automatically inferring
which argument in the recursive call gets smaller with respects to the original
input argument.  For productivity, the algorithm checks that the corecursive
call appears directly under a constructor to make sure that this function
always produces at least one element after each recursive step.
This implies that some functions, though well-defined, cannot be accepted by
the proof assistant. Examples of this include common sorting algorithms, such
as quicksort and mergesort.
%One example is Quicksort, written in OCaml
%below:
\begin{minted}{OCaml}
let rec qsort xs = match divide xs with | None -> []
 | Some (pivot, (smaller, larger)) -> qsort smaller @ (pivot::qsort larger)
\end{minted}
While \ocaml{qsort} is a well-defined mathematical function it cannot be
accepted by a proof assistant.  The reason is that the \ocaml{divide}
function \emph{destructuring} the input ``dives'' deeper in the input returning
two sublists with the head as a pivot.

@@

This work presents the first formalisation in Coq of \emph{hylomorphisms}, as
well as their algebraic laws that: (1) is \emph{axiom-free}; and (2)
allows the extraction of naturally-looking, clean OCaml code. While programmers
still need to reason about the termination of their programs, our mechanisation
will allow the use of the algebraic laws of hylomorphisms for program
reasoning, as well as extracting to clean, naturally-looking code. As an
example our mechanisation of \ocaml{qsort} as a hylomorphism will produce
the following code:
\begin{coqcode}
let rec qsort = function | [] -> []
  | h :: t -> let (l, r) = partition (fun x0 -> (<=) x0 h) t in
              let cx = fun p -> match p with | Lbranch -> l | Rbranch -> r in 
              app (qsort (cx Lbranch)) (h :: (qsort (cx Rbranch)))
\end{coqcode}
The contributions we make in this paper are as follows:
\begin{itemize}
  \item We provide a framework for \emph{generic programming} with recursion
    schemes in Coq, with a proof of their algebraic laws and 
    program equivalences, that allows clean code extraction.
  \item We formalise divide and conquer, dynamic programming, and
    mutual recursion algorithms.
  \item We verified the short-cut fusion optimisation in Coq and
    extract optimised code to OCaml.
\end{itemize}

@@

Thus, in general given an $F$-algebra and $F$-coalgebra, the hylomorphism is the
unique solution (when it exists) to the equation
\begin{equation}
  \label{eq:hylo}
  f = a \circ F\; f \circ c
\end{equation}
However this solution does not necessarily exist for an arbitrary pair of
algebras and coalgebras. In fact, this definition cannot be accepted by Coq,
since there is no structurally smaller argument to do recursion on and there is
no corecursive definition that is guaranteed to be syntactically guarded -- as
required by Coq -- due to the use of an arbitrary algebra $a$.
To address these issues, we mechanise \emph{recursive
hylomorphisms} which are guaranteed to have a unique solution to the
hylomorphism equation. These are hylomoprhisms where the coalgebra is
\emph{recursive}, i.e. coalgebras that terminate on all inputs. The following
predicate captures that a coalgebra terminates on an input: 
\begin{coqcode}
Inductive RecF (h : Coalg F A) : A -> Prop :=
| RecF_fold x : (forall e, RecF h (cont (h x) e)) -> RecF h x.
\end{coqcode}
A recursive coalgebra of type \coq{RCoalg F A} is a coalgebra 
\coq{c : Coalg F A} such that \coq{forall x, RecF c x}.
Recursive hylomorphisms are implemented in Coq recursively on the structure of
the proof for the type (\coq{RecF}) as follows:
\begin{coqcode}
Definition hylo_def (a : Alg F B) (c : Coalg F A) 
  : forall (x : A), RecF c x -> B := fix f x H
  := match c x as h0 return (forall e : Pos (shape h0), RecF c (cont h0 e)) -> B
     with | MkApp s_x c_x => fun H => a (MkApp s_x (fun e => f (c_x e) (H e)))
     end (RecF_inv H).
\end{coqcode}
We use \coq{RecF_inv} to obtain the structurally smaller proof to use in
the recursive calls. As we did with catamorphisms and anamorphisms, we
prove that \coq{hylo_def} is respectful, and use this proof to build the
corresponding higher-order proper morphism:
\begin{coqcode}
  hylo : forall `{F : Cont Sh P} `{setoid A} `{setoid B}, 
    Alg F B ~> RCoalg F A ~> A ~> B
\end{coqcode}
Finally, we show that recursive hylomorphisms are the unique solution to the
hylomorphism equation.
\begin{coqcode}
Lemma hylo_univ (g : Alg F B) (h : RCoalg F A) (f : A ~> B)
    : f =e hylo g h <-> f =e g \o fmap f \o h.
\end{coqcode}
Hylomorphisms fusion falls out from the universal property:
\begin{coqcode}
Lemma hylo_fusion_l (h1 : RCoalg F A) (g1 : Alg F B) (g2 : Alg F C)
  (f2 : B ~> C) (E2 : f2 \o g1 =e g2 \o fmap f2) : f2 \o hylo g1 h1 =e hylo g2 h1.
\end{coqcode}
% Qed.
%   rewrite hylo_univ, fmap_comp, compA, <- E2,
%           <- compA, <- compA, (compA g1), <- hylo_unr.  reflexivity.

@@

% Similarly to the rest of the definitions, we prove that recursive anamorphisms
% are respectful, and use this fact to build them as higher order proper
% morphisms, and prove their universal law, i.e.\ that recursive anamorphisms are
% the unique solution to the equation
% \coq{f =e l_in \o fmap f \o h},
% where \coq{h} is a recursive coalgebra.

\section{Extraction}
The steps are the following:
\begin{enumerate}
  \item Write a specification as a morphism \coq{f : A ~> B}.
  \item Use Coq's generalised rewriting and the laws of morphisms to 
    rewrite it to some \coq{g : A ~> B} such that \coq{f =e g}.
  \item Project the function \coq{h : A -> B} from \coq{g}
    (i.e. \coq{h = app g}).
  \item Extract \coq{h} to OCaml.
\end{enumerate}
Note that steps 2 and 3 are \emph{proofs} that \coq{f} and \coq{h} are
\emph{extensionally equal}. Then, Coq's extraction mechanism will erase any
proofs from the code, and inline our definitions. To ease the application of
these steps, we provide the following helper type:
\begin{coqcode}
  Record Ext (f : A ~> B) := 
    MkExt { target :> A -> B; tgt_eq : app f =e target; }.
\end{coqcode}
Note that \coq{tgt_eq} will be erased when extracting \coq{Ext f} to OCaml,
since it is in \coq{Prop}. This implies that the code extracted for \coq{Ext f}
will be the function \coq{target} that is extensionally equal to \coq{f}.  We
use a shortcut tactic, \coq{calculate}, that is a synonym to 
\coq{eapply MkExt}. As a result of this tactic, we will be given the goal 
\coq{f =e ?Goal0}. We can rewrite \coq{f}, unfold definitions, etc, and
complete the definition with \coq{reflexivity}, and this will instantiate the
metavariable \coq{?Goal0} to our desired function. We will use this extensively
in our examples.

\subsection{Sorting Algorithms}
So far, our examples could be modelled exclusively using catamorphisms. Here, we
show examples that require recursive hylomorphisms and termination proofs.  We
complete the sorting algorithm examples by applying fusion optimisation.

Both mergesort and quicksort are divide-and-conquer algorithms that can be
captured by the structure of an hylomorphisms. The structure of the recursion
that of a binary tree.  For example, in the case of quicksort, a list is split
into a pivot, the label of the node, and two sublists. We define the data
functor of trees as follows:
\begin{coqcode}
Inductive ITreeF L N X := i_leaf (l : L) | i_node (n : N) (l r : X)
\end{coqcode}
We define the functor as a container using the following shapes and positions:
\begin{coqcode}
Inductive Tshape L A := | Leaf (ELEM : L) | Node (ELEM : A).
Inductive Tpos := | Lbranch | Rbranch. 
\end{coqcode}
We define the tree container, \coq{TreeF}, in a straightforward way, by making
the positions of type \coq{Tpos} only valid in \coq{Node}.
We define a series of definitions for tree container constructors and 
destructors:
\mpav{The following algebra should be called a\_in or l\_in whatever}
\begin{coqcode}
Definition a_out {L A X : Type} : App (TreeF L A) X ~> ITreeF L A X.

(* Converting TreeF to ITreeF *)
Notation a_leaf x := (MkCont (Leaf _ x) (@dom_leaf _ _ _ x)).
Notation a_node x l r := (MkCont (Node _ x)
  (fun p => match val p with | Lbranch => l | Rbranch => r end)).
\end{coqcode}

% \subsubsection{Quicksort}

@@

Definition c_split : Coalg (TreeF unit int) (list int).
|{ x ~> match x with
        | nil => a_leaf tt
        | cons h t => let (l, r) := List.partition (fun x => x <=? h) t in
                      a_node h l r
        end}|.
Defined.
\end{coqcode}
We prove that the coalgebra \coq{c_split} is recursive by showing
that it respects the ``less-than'' relation on the length of the lists.
We then define, and extract, the hylomorphism \coq{hylo merge r_split}, which
extracts to the following OCaml code:
\begin{ocamlcode}
let rec qsort = function
| [] -> [] | h :: t ->
  let (l, r) = partition (fun x0 -> leb x0 h) t in
  let x0 = fun e -> qsort (match e with | Lbranch -> l | Rbranch -> r) in
  app (x0 Lbranch) (h :: (x0 Rbranch))
\end{ocamlcode}
Note that Coq's code extraction is unable to inline \coq{x0}, but the resulting
code is similar to a hand-written \coq{qsort}.

The mergesort algorithm can be defined analogously and can be found in the formalisation. [REF LINK].

% \subsubsection{Mergesort}

% Mergesort is a hylomorphism with the recursive structure defined by the
% container \coq{(TreeF (list int) unit)}.
% The computationally expensive part of the Mergesort implementation is the
% algebra, since this is where most of the comparisons happen. The recursive
% coalgebra for Mergesort is defined as the following Coq functions:
% \begin{coqcode}
% Fixpoint splitL (x : list int) (accL accR : list int) :=
%   match x with
%   | nil => (accL, accR)
%   | cons x xs => splitL xs accR (cons x accL)
%   end.

% Definition c_split : Coalg (TreeF (list int) unit) (list int) :=
% |{ x ~> match x with
%         | nil | cons _ nil => a_leaf x
%         | _ => let (l, r) := splitL x nil nil in
%                a_node tt l r
%         end
%  }|.
% Defined.
% \end{coqcode}
% Here, the notation \coq{|{ x ~> E }|} defines the morphism \coq{fun x => E},
% and uses a custom tactic to prove that it is respectful. The recursive
% coalgebra \coq{c_split} takes a list, and it returns a leaf with the input
% list, if its size is less or equal than 1. Otherwise, it applies the function
% \coq{splitL}, and divides the list in two halves.

% It is easy to prove that this coalgebra is recursive, by showing that
% it respects the relation \coq{<} on the length of the lists:
% \begin{coqcode}
% Lemma split_fin : respects_relation c_split (@length int) lt.
%   Proof. (* *) Qed.

% Definition tsplit : RCoalg (TreeF (list int) unit) (list int)
%   := mk_wf_coalg wf_lt split_fin.
% \end{coqcode}

% The algebra simply applies the \coq{mergeL} function in order to the sublists
% in every node:
% \begin{coqcode}
% Fixpoint mergeL l1 l2 {struct l1}  : list int := (* *)

% Definition merge : App (TreeF (list int) unit) (list int) ~> list int.
% |{ x ~> match a_out x with | node _ l r => mergeL l r | leaf e => e end }|.
% Defined.
% \end{coqcode}

% Putting everything together, Mergesort is defined as the following
% hylomorphism:
% \begin{coqcode}
% Definition msort := hylo merge tsplit.
% \end{coqcode}
% Note, however, that we could have composed the catamorphism and anamorphism,
% and used the deforestation result to optimise the intermediate binary tree
% away:
% \begin{coqcode}
% Definition msort : Ext (cata merge \o rana tsplit).
% calculate. rewrite (deforest l_in_out). simpl. reflexivity. Defined.
% \end{coqcode}
% In both cases, the result is the same, and it extracts to the usual divide and
% conquer definition of Mergesort in OCaml:
% \begin{ocamlcode}
% let rec msort x = match x with
% | [] -> x | _ :: l -> (match l with | [] -> x | _ :: _ ->
%      let (l0, r) = splitL x [] [] in
%      let cx = fun p -> match p with | Lbranch -> l0 | Rbranch -> r in
%      mergeL (msort (cx Lbranch)) (msort (cx Rbranch)))
% \end{ocamlcode}
\subsubsection{Fusing a divide-and-conquer computation}
